Bandit Multiclass Linear Classification: Efficient Algorithms for the Separable Case

### 2019-01-23 ICML-submission/introduction.tex
Fixes

Finally, we study two questions related to the computational and
information-theoretic hardness of the problem. Any algorithm for the bandit
setting collects information in the form of so called \emph{strongly labeled}
and \emph{weakly labeled} examples. Strongly labeled examples are those for
which we know the class label. Weakly labeled example is an example for which we
know that class label can be anything except for a particular one class.
In Appendix~\ref{section:np-hardness-of-weak-labeling-problem}, we show that the offline
problem of finding a linear multiclass separator consistent with a set of
strongly and weakly labeled examples is NP-hard. In
Appendix~\ref{section:mistake-lower-bound-for-ignorant-algorithms}, we show that
any algorithm that uses only strongly-labeled examples and ignores weakly
labeled examples makes at least $\Omega\left( \left(1/(160 \gamma)\right)^{(d-2)/4} \right)$
mistakes.

>>>>

Finally, we study two questions related to the computational and
information-theoretic hardness of the problem. Any algorithm for the bandit
setting collects information in the form of so called \emph{strongly labeled}
and \emph{weakly labeled} examples. Strongly labeled examples are those for
which we know the class label. Weakly labeled example is an example for which we
know that class label can be anything except for a particular one class. In
Appendix~\ref{section:np-hardness-of-weak-labeling-problem}, we show that the
offline problem of finding a linear multiclass separator consistent with a set
of strongly and weakly labeled examples is NP-hard. In
Appendix~\ref{section:mistake-lower-bound-for-ignorant-algorithms}, we prove a
lower bound on the number of mistakes of any algorithm that uses only
strongly-labeled examples and ignores weakly labeled examples.

### 2019-01-23 ICML-submission/mistake-lower-bound-for-ignorant-algorithms.tex
Fixes

The ignorant algorithm is formally defined as follows.
\begin{definition}
\label{definition:ignorant-algorithm}
A bandit classification algorithm $\calA$ is ignorant if for all $t \in
\{1,2,\dots,T\}$, $\widehat y_t\in \{1,2\}$ is drawn from some distribution that
is jointly determined by
\begin{enumerate}
\item $x_t$
\item $H_t \triangleq \left( (x_{\tau_1}, y_{\tau_1}), (x_{\tau_2}, y_{\tau_2}), \ldots, (x_{\tau_{n_t}}, y_{\tau_{n_t}}) \right)$,
where $1 \le \tau_1 < \dots < \tau_{n_t} < t$, and $\{\tau_1, \ldots, \tau_{n_t}\}$
are the rounds in which the algorithm predicts correctly (that is,
$\{\tau_1, \ldots, \tau_{n_t}\} = \{1 \le \tau < t ~:~ \widehat y_\tau=y_\tau\}$).
\end{enumerate}
We use $p_t(\cdot|x_t)\in \{p\in \R_+^2 ~:~ p^\top\mathbf{1}=1\}$ to denote
the distribution $\widehat y_t$ is drawn from. Note that we let the current
feature vector $x_t$ be an argument to $p_t$ to incorporate its dependency on
$x_t$. Therefore, for an ignorant algorithm, the function $p_t(\cdot|\cdot)$ is
only determined by $H_t$.
\end{definition}

An ignorant algorithm does not learn from mistakes in the following sense.
Suppose that an ignorant algorithm predicts incorrectly at time $t$ ($\widehat
y_t \neq y_t$). Then at time $t+1$, if it is given the same feature vector
$x_{t+1}=x_t$, it will predict the label with the same distribution
$p_{t+1}(\cdot|x_{t+1})=p_{t+1}(\cdot|x_t)=p_t(\cdot|x_t)$. Below we describe
the difficulty of learning from mistakes using a class of natural approaches
under the weak separability assumption. Consider an approach similar to OBAMA:
in each round $t$, we create a \textit{convex} loss function $\widehat \ell_t:
\R^{K\times d} \to \R$ that gives a loss value $\widehat
\ell_t(W)$ to all linear models $W \in \R^{K\times d}$ ($W=[w_1 \ldots
w_K]^\top$ for $w_1, \ldots, w_K$ defined in
Definition~\ref{definition:linear-separability}). The requirement for
$\widehat{\ell}_t$ to be convex is due to computational efficiency. To let
$\widehat{\ell}_t$ reflect the loss, if $\widehat y_t \neq y_t$, we want to
assign larger loss values $\widehat{\ell_t}(W)$ to the inconsistent models
$\mathcal{W}=\{W: \ip{w_{\widehat y_t}}{x_t} \ge \ip{w_i}{x_t}, \forall
i\in[K]\}$, while for other $W$'s, we want $\widehat{\ell_t}(W)$ to be smaller.
The following fact can be verified when $K\ge 3$, which suggests there is no
way to define the loss $\widehat \ell_t$ that satisfies the above properties
constraints: there exists $W_1, W_2, W_3$, where $W_2$ is in the line segment
between $W_1$ and $W_3$, such that $W_2\in \mathcal{W}$, while $W_1, W_3 \notin
\mathcal{W}$. On one hand, by convexity, we should have
$\widehat{\ell}_t(W_2)\le \max\{\widehat{\ell}_t(W_1),
\widehat{\ell}_t(W_3)\}$; on the other hand, to penalize $\mathcal{W}$, we
should have $\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_1)$ and
$\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_3)$. Clearly, these two requirements
contradict each other.

@@

\begin{theorem}[Mistake Lower Bound for Ignorant Algorithms]
\label{theorem:ignorant_lower_bound}
Suppose $\calA$ is an ignorant bandit classification algorithm. There
exists an adversary that can sequentially choose $(x_1, y_1), (x_2, y_2), \dots,
(x_T, y_T) \in \R^d\times \{1,2\}$ so that the examples $\gamma$-strongly
linearly separable, $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$, and the
expected number of mistakes made by $\calA$ is at least
$$
\frac{1}{10}\min\left\{\sqrt{T}, \left(\frac{1}{160\gamma}\right)^{\frac{d-2}{4}} \right\} \; .
$$
\end{theorem}

@@

\begin{lemma}
\label{lemma:embed_d_gamma}
Let $\gamma \in (0,\frac 1 {160})$, let $d$ be a positive integer and let $N = (\Omega(\frac{1}{\gamma}))^{(d-2)/2}$.
There exist vectors $u_1, u_2, \dots, u_N, v_1, v_2, \dots, v_N \in \R^d$ such that for all $i, j \in \{1,2,\dots,N\}$,
\begin{align*}
\norm{u_i} & \le 1 \; , \\
\norm{v_j} & \le 1 \; , \\
\ip{u_i}{v_j} & \ge \gamma & & \text{if $i=j$,} \\
\ip{u_i}{v_j} & \le -\gamma & & \text{if $i \neq j$.}
\end{align*}
\end{lemma}

\begin{proof}
By Lemma 6 of~\cite{Long-1995}, there exists a set of $N =
(\frac{1}{2\sqrt{40\gamma}})^{d-2}$ $(d-1)$-dimensional vectors $z_1, \ldots,
z_N$ on the unit sphere $\mathbb{S}^{d-2}$, such that $\theta(z_i, z_j) \ge
\sqrt{40 \gamma}$. As $\cos\theta \le 1-\theta^2/5$ for every $\theta \in
[0,\pi]$, this implies that
\begin{align*}
\ip{z_i}{z_j} &= 1 && \text{if $i = j$,} \\
\ip{z_i}{z_j} &\le 1 - 8\gamma && \text{if $i \neq j$.}
\end{align*}

@@

\begin{proof}[Proof for Theorem~\ref{theorem:ignorant_lower_bound} (Part I: Mistake lower bound)]
We consider the strategy for the adversary described in Algorithm~\ref{algorithm:adversary-strategy}.
\begin{algorithm}[h]
\caption{\textsc{Adversary's strategy}}
\label{algorithm:adversary-strategy}
\begin{algorithmic}[1]
{
\STATE{\textbf{Define} $q_0=\max\left\{\frac{1}{\sqrt{T}}, \frac{1}{\sqrt{N}}\right\}$, where $N$ is defined in Lemma~\ref{lemma:embed_d_gamma}. Also, let $v_1, \ldots, v_N$ be those defined in Lemma~\ref{lemma:embed_d_gamma}.}
\STATE{\textbf{Initialize} $\textsc{phase}= 1$. }
\FOR{$t=1,\dots,\min\{T,N\}$}
    \IF{$\textsc{phase}=1$}
       \IF{$p_t(1|v_t)\ge 1-q_0$}
          \STATE{$(x_t, y_t)\leftarrow (v_t, 1)$}
       \ELSE
          \STATE{$(x_t, y_t)\leftarrow (v_t, 2)$}
          \STATE $\textsc{phase}\leftarrow 2$
       \ENDIF
    \ELSIF{$\textsc{phase}=2$}
       \STATE $(x_t, y_t)\leftarrow (x_{t-1}, y_{t-1})$.
    \ENDIF
\ENDFOR
\FOR{$t=\min\{T,N\}+1, \ldots, T$}
    \STATE $(x_t, y_t)\leftarrow (x_{t-1}, y_{t-1})$.
\ENDFOR

@@

\begin{proof}[Proof for Theorem~\ref{theorem:ignorant_lower_bound} (Part II: linear separability)]
Observe that in Algorithm~\ref{algorithm:adversary-strategy}, in the beginning
the adversary always set the label to $1$; once it sets $y_t=2$ at some $t$, it
lets $(x_s, y_s)=(x_t, y_t)=(v_t, 2)$ for all $s\ge t$. Consider $w_2=u_t$ and
$w_1=-u_t$, for $u_t$ as defined in Lemma~\ref{lemma:embed_d_gamma}. Then by
Lemma~\ref{lemma:embed_d_gamma}, we have $\ip{w_2}{x_s} = \ip{u_t}{v_t} \ge
\gamma, \forall s \ge t$ and $\ip{w_2}{x_s} = \ip{u_t}{v_s} \le - \gamma,
\forall s< t$; also $\ip{w_1}{x_s} \le -\gamma, \forall s \ge t$ and
$\ip{w_1}{x_s} \ge  \gamma, \forall s<t$. Thus, the example set is always
$\gamma$-strongly separable.
\end{proof}

>>>>

The ignorant algorithm is formally defined as follows.
\begin{definition}
\label{definition:ignorant-algorithm}
An algorithm $\calA$ for \textsc{Online Multiclass Linear Classification with
Bandit Feedback} is called \emph{ignorant} if for all $t \in \{1,2,\dots,T\}$,
$\widehat y_t$ is drawn from some distribution that is jointly determined by
\begin{enumerate}
\item $x_t$
\item $H_t \triangleq \left( (x_{\tau_1}, y_{\tau_1}), (x_{\tau_2}, y_{\tau_2}), \ldots, (x_{\tau_{n_t}}, y_{\tau_{n_t}}) \right)$,
where $1 \le \tau_1 < \dots < \tau_{n_t} < t$, and $\{\tau_1, \ldots, \tau_{n_t}\}$
are the rounds in which the algorithm predicts correctly (that is,
$\{\tau_1, \dots, \tau_{n_t}\} = \{1 \le \tau < t ~:~ \widehat y_\tau=y_\tau\}$).
\end{enumerate}
We use $p_t(\cdot|x_t)\in \{p\in \R_+^K ~:~ p^\top\mathbf{1}=1\}$ to denote
the distribution $\widehat y_t$ is drawn from. Note that we let the current
feature vector $x_t$ be an argument to $p_t$ to incorporate its dependency on
$x_t$. Therefore, for an ignorant algorithm, the function $p_t(\cdot|\cdot)$ is
only determined by $H_t$.
\end{definition}

An ignorant algorithm does not learn from mistakes in the following sense.
Suppose that an ignorant algorithm predicts incorrectly at time $t$. Then at
time $t+1$, if it is given the same feature vector $x_{t+1}=x_t$, it will
predict the label with the same distribution
$p_{t+1}(\cdot|x_{t+1})=p_{t+1}(\cdot|x_t)=p_t(\cdot|x_t)$.

% Below we describe
%the difficulty of learning from mistakes using a class of natural approaches
%under the weak separability assumption. Consider an approach similar to OBAMA:
%in each round $t$, we create a \textit{convex} loss function $\widehat \ell_t:
%\R^{K\times d} \to \R$ that gives a loss value $\widehat
%\ell_t(W)$ to all linear models $W \in \R^{K\times d}$ ($W=[w_1 \dots
%w_K]^\top$ for $w_1, w_2, \dots, w_K$ defined in
%Definition~\ref{definition:linear-separability}). The requirement for
%$\widehat{\ell}_t$ to be convex is due to computational efficiency. To let
%$\widehat{\ell}_t$ reflect the loss, if $\widehat y_t \neq y_t$, we want to
%assign larger loss values $\widehat{\ell_t}(W)$ to the inconsistent models
%$\mathcal{W}=\{W: \ip{w_{\widehat y_t}}{x_t} \ge \ip{w_i}{x_t}, \forall
%i\in[K]\}$, while for other $W$'s, we want $\widehat{\ell_t}(W)$ to be smaller.
%The following fact can be verified when $K\ge 3$, which suggests there is no
%way to define the loss $\widehat \ell_t$ that satisfies the above properties
%constraints: there exists $W_1, W_2, W_3$, where $W_2$ is in the line segment
%between $W_1$ and $W_3$, such that $W_2\in \mathcal{W}$, while $W_1, W_3 \notin
%\mathcal{W}$. On one hand, by convexity, we should have
%$\widehat{\ell}_t(W_2)\le \max\{\widehat{\ell}_t(W_1),
%\widehat{\ell}_t(W_3)\}$; on the other hand, to penalize $\mathcal{W}$, we
%should have $\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_1)$ and
%$\widehat{\ell}_t(W_2) > \widehat{\ell}_t(W_3)$. Clearly, these two requirements
%contradict each other.

@@

\begin{theorem}[Mistake Lower Bound for Ignorant Algorithms]
\label{theorem:ignorant-lower-bound}
Suppose $\calA$ is an ignorant algorithm for \textsc{Online Multiclass Linear
Classification with Bandit Feedback}. There exists an adversary that can
sequentially choose labeled examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)
\in \R^d\times \{1,2\}$ so that the examples are $\gamma$-strongly linearly
separable, $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$, and the expected
number of mistakes made by $\calA$ is at least
$$
\frac{1}{10}\min\left\{\sqrt{T}, \left(\frac{1}{160\gamma}\right)^{\frac{d-2}{4}} \right\} \; .
$$
\end{theorem}

@@

\begin{lemma}
\label{lemma:embed_d_gamma}
Let $\gamma \in (0,\frac{1}{160})$, let $d$ be a positive integer and let $N = (\frac{1}{2\sqrt{40\gamma}})^{d-2}$.
There exist vectors $u_1, u_2, \dots, u_N, v_1, v_2, \dots, v_N \in \R^d$ such that for all $i, j \in \{1,2,\dots,N\}$,
\begin{align*}
\norm{u_i} & \le 1 \; , \\
\norm{v_j} & \le 1 \; , \\
\ip{u_i}{v_j} & \ge \gamma & & \text{if $i=j$,} \\
\ip{u_i}{v_j} & \le -\gamma & & \text{if $i \neq j$.}
\end{align*}
\end{lemma}

\begin{proof}
By Lemma 6 of~\citet{Long-1995}, there exists a set of $N$ dimensional vectors
$z_1, z_2, \dots, z_N \in \R^{d-1}$ such that $\norm{z_1} = \norm{z_2} = \dots =
\norm{z_N} = 1$ and the angle between the vectors is $\measuredangle(z_i, z_j) \ge
\sqrt{40 \gamma}$. Since $\cos\theta \le 1-\theta^2/5$ for any $\theta \in
[-\pi,\pi]$, this implies that
\begin{align*}
\ip{z_i}{z_j} &= 1 && \text{if $i = j$,} \\
\ip{z_i}{z_j} &\le 1 - 8\gamma && \text{if $i \neq j$.}
\end{align*}

@@

\begin{proof}[Proof for Theorem~\ref{theorem:ignorant-lower-bound} (Part I: Mistake lower bound)]
We consider the strategy for the adversary described in Algorithm~\ref{algorithm:adversary-strategy}.
\begin{algorithm}[h]
\caption{\textsc{Adversary's strategy}}
\label{algorithm:adversary-strategy}
\begin{algorithmic}[1]
{
\STATE{\textbf{Define} $q_0=\max\left\{\frac{1}{\sqrt{T}}, \frac{1}{\sqrt{N}}\right\}$, where $N$ is defined in Lemma~\ref{lemma:embed_d_gamma}. Also, let $v_1, \ldots, v_N$ be those defined in Lemma~\ref{lemma:embed_d_gamma}.}
\STATE{\textbf{Initialize} $\textsc{phase}= 1$. }
\FOR{$t=1,\dots,\min\{T,N\}$}
    \IF{$\textsc{phase}=1$}
       \IF{$p_t(1|v_t)\ge 1-q_0$}
          \STATE{$(x_t, y_t)\leftarrow (v_t, 1)$}
       \ELSE
          \STATE{$(x_t, y_t)\leftarrow (v_t, 2)$}
          \STATE $\textsc{phase}\leftarrow 2$
       \ENDIF
    \ELSIF{$\textsc{phase}=2$}
       \STATE $(x_t, y_t)\leftarrow (x_{t-1}, y_{t-1})$.
    \ENDIF
\ENDFOR
\FOR{$t=\min\{T,N\}+1, \ldots, T$}
    \STATE $(x_t, y_t)\leftarrow (x_{t-1}, y_{t-1})$.
\ENDFOR

@@

\begin{proof}[Proof for Theorem~\ref{theorem:ignorant-lower-bound} (Part II: linear separability)]
Observe that in Algorithm~\ref{algorithm:adversary-strategy}, in the beginning
the adversary always set the label to $1$; once it sets $y_t=2$ at some $t$, it
lets $(x_s, y_s)=(x_t, y_t)=(v_t, 2)$ for all $s\ge t$. Consider $w_2=u_t$ and
$w_1=-u_t$, for $u_t$ as defined in Lemma~\ref{lemma:embed_d_gamma}. Then by
Lemma~\ref{lemma:embed_d_gamma}, we have $\ip{w_2}{x_s} = \ip{u_t}{v_t} \ge
\gamma, \forall s \ge t$ and $\ip{w_2}{x_s} = \ip{u_t}{v_s} \le - \gamma,
\forall s< t$; also $\ip{w_1}{x_s} \le -\gamma, \forall s \ge t$ and
$\ip{w_1}{x_s} \ge  \gamma, \forall s<t$. Thus, the example set is always
$\gamma$-strongly separable.
\end{proof}
